--- 
title: "Introduction to Spatial Analysis in R"
author: "D March, J Duffy & Chris Yeomans"
site: bookdown::bookdown_site
output:
    bookdown::gitbook:
        config:
            sharing: null
        css: 'style.css'
        includes:
            in_header: _toggle.html
        keep_md: TRUE
linkcolor: blue
documentclass: book
link-citations: yes
description: "Skeleton Tutorial Template"
---

```{r, child = "_setup.Rmd", include = F, purl = F, cache = F}
```

# Introduction

In this workshop we are going to work with vector data (points, lines and polygons) and raster data. R has an extraordinary potential to handle this data and perform complex spatial operations like other Geographic Information Systems (GIS) software.

We will explore the capacities of R to handle and manipulate vector data through two worked examples:

1. Hello, world!

2. Animal oceanographers

We will also play with raster data with a further two examples: 

1. Introduction to raster data

2. Working with real world data


Accompanying slides can be downloaded from [here](slides.zip)

## Set up your working environment

1. Create a new project within Rstudio. If you are unsure, instructions on how to do this can be found [here](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects).

2. Download the [Data](data.zip).

3. Unzip and store its contents in a folder named `data` within your new project.

4. Now create the following folders within your project - `output` & `img`. This can be done from within R as follows:

```{r create_folders, warning = FALSE}
dir.create("output")
dir.create("img")
```

5. Run the following code to install all required packages:

```{r message = FALSE, warning = FALSE, eval = FALSE}
install.packages(c("rgdal",
                 "raster",
                 "viridis",
                 "leaflet",
                 "ggplot2",
                 "mapproj",
                 "robis",
                 "gifski",
                 "elevatr",
                 "move",
                 "moveVis",
                 "dplyr",
                 "lubridate",
                 "rgeos",
                 "adehabitatHR",
                 "RColorBrewer",
                 "rasterVis",
                 "rgl"))
```

5. Once your working environment is set up, you can copy/paste the lines of code from this website into R script(s) to run.
